R Tools for ObsPack, Receptors and Footprints (rtorf) for processing atmospheric observations in NOAA-GML

### 2024-08-27 paper/paper.Rmd
add paper
paper

---
title: "robspack"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

![GitHub commit
activity](https://img.shields.io/github/commit-activity/y/ibarraespinosa/robspack)
[![R build
status](https://github.com/ibarraespinosa/robspack/workflows/R-CMD-check/badge.svg)](https://github.com/ibarraespinosa/robspack/actions)
[![R build
status](https://github.com/ibarraespinosa/robspack/workflows/draft-pdf/badge.svg)](https://github.com/ibarraespinosa/robspack/actions)

[NOAA Obspack](https://gml.noaa.gov/ccgg/obspack/) is a collection of green house gases observations

`robspack` only depends on `data.table`, which is basically parallel C, 
so it can be installed in any machine.

## installation

```{r, eval = F}
remotes::install_github("ibarraespinosa/robspack")
```

```{r, include = T, message=F, warning=F}
library(robspack)
library(data.table)
```

## ObsPack summary

The first step consists in constructing a summary for the ObsPack.
This is required to read the data, but also, identify `agl`, 
which is present in some of the file names.
This function returns a `data.frame`.
Optionally, the user can indicate a path to store the `data.frame`.
`obs_summary` also prints a summary of the data. The second
argument is the categories, and by default includes the categories
shown below, to account for all the files.
Then the summary `data.frame` contains 
the columns `id` as the full path to each
file, `name` which is the name or relative path of the file, `n`
just an id, `sector` such as tower, and the column `agl`
which indicates the `agl` indicated in the name of the file
if available. To read the documentation of this function,
the user must run `?obs_summary`.

```{r}

categories <- c(
  "aircraft-pfp",
  "aircraft-insitu",
  "surface-insitu",
  "aircore",
  "surface-pfp",
  "tower-insitu",
  "shipboard-insitu",
  "flask"
)
obs <- "../../obspack_ch4_1_GLOBALVIEWplus_v4.0_2021-10-14/data/txt"
index <- obs_summary(obs = obs)

```

There are 362 files in the ObsPack directory.
The printed information also shows the total at the bottom,
as the sum of the individual file by sector. This
is to ensure that the sum of files is equal to the
total number of files found, shown at the top.
furthermore, the printed information
also shows that there are 136 files with the
`agl` explicitly mentioned in the name of the file.

Sometimes we need more information about the site. For instance,
what do the observations start and end. Then, we added 
the function `obs_table`, which calculates statistics
summary of "time" and other numeric variables by file name,
sector, site, altitude and mode. For instance,
the observations in the site "SCT" in
South Carolina, USA, were between "2015-08-19 21:30:00 UTC" and
"2020-12-31 23:30:00 UTC".
```{r}
dft <- obs_table(index = index,
                 categories = "tower-insitu",
                 verbose = FALSE)
dft[site_code == "SCT", ]$timeUTC |> 
  range()
```

### Read data

Once the summary is built, the function `obs_read`
will read the files available in the
index file previously generated.
Here we selected the category "tower-insitu".
The argument verbose prints which files are being read each time,
by default.
At the end, this function prints the total number
of observations by type of altitude (agl or asl).

```{r}
df <- obs_read(index = index,
               categories = "tower-insitu",
               verbose = FALSE)

```

We added a function to plot the data read from ObsPack.
The y-axis is the field `value` and the 
x-axis is by default `time`. The data illustrated sorted by
color is the field `site_code`, with the default
number of 3 sites. The argument `pal` is to define the
color palette, used by the internally imported function
`cptcity::cpt`.

```{r, tidy = TRUE, dpi=300, fig.cap="First two sites in ObsPack"}
obs_plot(dt = df, time = "time", yfactor = 1e9, cex = 0.5)
```

Here we can see `r round(nrow(df)/1000000, 2)` million observations 
for `tower-insitu`. These observations are made between 
`r range(df$year)[1]` and `r range(df$year)[2]`. The identification
of the altitude and type is critical. The approach used here 
consists of:

1. Identify `agl` from the name of the tile.
2. If `agl` not present, search fill_values used in elevation and transform them into NA (not available)
3. If `agl` is not present, `agl = altitude - elevation`.
4. If there are some NA in elevation, will result some NA in `agl`
5. A new column is added named `altitude_final` to store `agl` or `asl`
6. Another column named `type_altitude` is added to identify `agl` or `asl`.
7. If there is any case NA in `altitude_final`, `type_altitude` is "not available"

### Filtering

ObsPack includes global observations and sometimes we need to extract data
for a specific region and periods of time.
In this part we include spatial and temporal parameters
to filter data.
The year of interest is 2020, but we also included
December of 2019 and January of 2021.
At this stage, we can apply the spatial filter
by using the coordinates.

```{r}
north <- 80
south <- 10
west <- -170
east <- -50
max_altitude <- 8000
evening <- 14:15

yy <- 2020
df <- rbind(df[year == yy - 1 & month == 12],
            df[year == yy],
            df[year == yy + 1 & month == 1])

df <- df[altitude_final < max_altitude &
           latitude < north &
           latitude > south &
           longitude < east &
           longitude > west]
```

After filtering by space and time, we have `r round(nrow(df), 2)`
million observations. Towers can have observations at different
heights. Here we need to select one site with the observations
registered at the highest height. The column
with the height is named `altitude_final` and the
max altitude was named `max_altitude`.

```{r}
dfa <- df[,
          max(altitude_final),
          by = site_code] |> unique()

names(dfa)[2] <- "max_altitude"
```

### Time

Here we need to start time columns.
The function `obs_addtime` adds time columns
`timeUTC`, `timeUTC_start` which shows the start time
of each observation and `timeUTC_end`
which shows the end time for each observation.
Then we need to identify the local time with the function
`add_ltime`. This is important because to identifying
observations in the evening in local time.
`add_ltime` uses two methods, first identify
the time difference with utc by identifying the
metadata column "site_utc2lst". If this information
is not available, with the aircrafts for instance,
the local time is calculated with an approximation
based on longitude:

$$
lt = UTC + longitude/15 * 60 * 60
$$
Where $lt$ is the local time, $UTC$ the time,
$longitude$ the coordinate.
Then, the time is cut every two hours.
Now, we identify the local time to select evening hours.

```{r}
df2 <- obs_addtime(df)
df2$timeUTC <- cut(x = df2$timeUTC+3600,
                   breaks = "2 hour") |>
  as.character() |>
  as.POSIXct(tz = "UTC")
df3 <- obs_addltime(df2)
df3 <- df3[lh %in% evening]
```

Now there are `r nrow(df3)` observations.
At this point we can calculate the averages of several columns by
the cut time. The function `obs_agg` does this aggregation as shown
in the following lines of code. The argument `gby`
establish the function used to aggregate `cols`, in this case
the function mean by time and altitude. Finally, we
add local time again.

```{r}
df4 <- obs_agg(dt = df3,
               gby = "mean",
               cols = c("value",
                        "latitude",
                        "longitude",
                        "type_altitude",
                        "dif_time",
                        "year_end",
                        "site_utc2lst"),
               verbose = FALSE,
               byalt = TRUE)
df5 <- obs_addltime(df4)
```

Now there are `r nrow(df5)` observations,
`r nrow(df3) - nrow(df5)` less observations.
Here we add the column `max_altitude` to identify the max altitude
by site.

```{r}
df5[,
    max_altitude := max(altitude_final),
    by = site_code]
df5[,
    c("site_code",
      "altitude_final",
      "max_altitude")] |> unique()
```

### Saving master as text and csvy

Now that we have all the required information, 
we can save the files. Here, we name the data.frame 
as master, because it contains all the information.
This is important because some 
fields can be used in the future, and for traceability.
For convenience,
time variables are transformed into character
before writing into the disk.
The separation is space " ".

```{r, eval = T}
tmp <- tempfile()
master <- df5
master$timeUTC <- as.character(master$timeUTC)
master$timeUTC_end <- as.character(master$timeUTC_end)
master$local_time <- as.character(master$local_time)

fwrite(master, 
       file =  "paper/tower_insitu_2020.txt",
       sep = " ")
```

The format Comma Separated Value with YAML (CSVY)[^3]
consists in a typical CSV with a YAML header.
The function`obs_write_csvy` includes the argument
`notes` which allows adding custom notes at the header of the file. Below the notes, `obs_write_csvy` adds 
the output of the R function `str`, which
provides a vertical summary of the data, known as
structure.

```{r, eval = T}
obs_write_csvy(dt = master,
          notes = "tower 2020",
          out = "paper/tower_insitu_2020.csvy")
```

To check the YAML header we read the first 38 lines of 
the files that were generated. Here we can see
the column names, type of data and first observations.
The YAML header is delimited by the characters "---".

```{r, eval = F}
readLines("paper/tower_insitu_2020.csvy")[1:38]
```

### Saving receptors

We need to filter some columns
from the master files in a new object called receptors.
This is needed because internally we run HYSPLIT [@hy]
using the information from the receptors.
In the case of a tower, we need to select 
observations with the highest 
altitude. The specific columns are selected
as shown on the following code. We are selecting the 
ending times, because later HYSPLIT is run backwards
based on the time of measurement, between ending and
starting times. The columns about time are formatted to
have two characters. For instance, the month 1,
is formatted as "01". We also need to filter for
`type_altitude` equal 0, representing `agl `observations
, or equal to 1, `asl`.

```{r, eval = F}

receptor <- master[altitude_final == max_altitude,
                   c("site_code",
                     "year",
                     "month",
                     "day",
                     "hour",
                     "minute",
                     "second",
                     "latitude",
                     "longitude",
                     "altitude_final",
                     "type_altitude",
                     "year_end",
                     "month_end",
                     "day_end",
                     "hour_end",
                     "minute_end",
                     "second_end")]
receptor$altitude_final <- round(receptor$altitude_final)
receptor <- obs_format(receptor)

if(nrow(receptor_agl) > 0) {
  fwrite(x = receptor_agl,
         file = "paper/receptor_tower_insitu_2020_AGL.txt"),
  sep = " ")}

if(nrow(receptor_asl) > 0) {
  fwrite(x = receptor_asl,
         file = "paper/receptor_tower_insitu_2020_ASL.txt"),
  sep = " ")}

```

## Recommendation for other applications
The approach to generate receptors depends on each type of 
observation and other considerations. For instance, aircraft with 
continuous observations at each second can be filtered and averaged 
every 20 seconds. In that way, the footprints are still 
representative and it would not be necessary to run HYSPLIT 
every second. Of course, it depends on the application 
and objective of the study. For this manuscript, we are 
presenting the generation of receptors based on tower observations.

## Application for other sectors
In this package we are sharing scripts to process other sectors
The scripts are available in the path 
`https://github.com/ibarraespinosa/robspack/tree/main/rscripts`

>>>>

---
# Example from https://joss.readthedocs.io/en/latest/submitting.html
title: "R Tools for ObsPack, Receptors and Footprints (rtorf) for processing atmospheric observations in NOAA-GML
' ObsPack"
tags:
  - R
  - GHG
  - NOAA
  - ObsPack
authors:
  - name: Sergio Ibarra-Espinosa
    orcid: 0000-0002-3162-1905
    affiliation: "1, 2" # (Multiple affiliations must be quoted)
  - name: Lei Hu
    orcid: 0000-0002-5624-3832
    affiliation: 2
affiliations:
 - name: Cooperative Institute for Research in Environmental Sciences, CU Boulder
   index: 1
 - name: NOAA Global Monitoring Laboratory
   index: 2
citation_author: Ibarra-Espinosa and Hu
date: 26 August 2024
year: 2024
bibliography: mybibfile.bib
output: rticles::joss_article
csl: apa.csl
journal: JOSS
---

# Summary

In this study, we present a new open-source R package `rtorf`, 
to read, process, select, and plot NOAA Observation Package (ObsPack)
data products. We use a methane ObsPack data product as an example
in this code base, but it can be easily modified to analyze ObsPack
products for other greenhouse gasses. The R package starts with 
creating a catalog of  all ObsPack files in each product. It then 
reads all files and creates one database. While reading each ObsPack 
file, it extracts site elevation and time zone information from the 
file header and calculates sampling altitude in meters above ground 
level and local time for individual samping events. Finally, it 
processes and selects observations for inverse modeling purposes. 
This package imports functions from data.table R package, which 
contains C bindings with parallel implementation via Open-MP 
[@dt]. rtorf provides functions 
to perform these tasks in a transparent and efficient way, 
supporting open-source communities in environmental sciences.

The world is experiencing an accelerated global warming
due to the accumulation of greenhouse gases (GHG)
since the industrial revolution [@us2018].
Greenhouse gas observations are critical to monitor the state of the
atmosphere, quantify present and historical emissions, and understand 
global climate change. 
During the 21th Conference of Parties (COP21), it was established the Paris 
Accord, a multilateral effort reduce greenhouse emissions
in order to limit the temperature increment of 1.5 degrees
[@rhodes20162015]. 
Methane is a greenhouse gas responsible for half of the 
temperature increase since preindustrial levels.
Furthermore, methane has a 9 years lifetime and a
global warming potential of 30 over 100 years [@epagwp], with
a current global radiative forcing of 0.650 $Wm^{-2}$ [@aggi].
Hence, in the 26 version of COP conference 
[@hunter2021glasgow], it was 
signed the Global Methane Pledge aiming reduce at least 
methane emissions 30\% from 2020 levels by 2030, with
U.S. as one of the parties [@wh]. Therefore, monitoring $CH_4$ observations,
emissions and sinks has become critical. NOAA 
ObsPack data has been used to 
support many studies. For instance, 
the global methane budget for the year 2017 was 596 $Tgy^{-1}$,
in agreement with other studies
[@saunois2020global;@saunois2016global]
@lu2021global, characterized global methane emissions in 
between 2014 and 2017, including a comparison with Greenhouse gases 
Observing SATellite (GOSAT) data. 
@saunois2016global.
At regional scale, @lu2022methane performed another studied focused
on north america using as priors local emissions inventories.

<!-- 2. Importance of greenhouse gases observations -->
<!-- 3. NOAA - mandate by congress - Obspack -->

The National Oceanic and Atmospheric Administration (NOAA) and its
Global Monitoring Laboratory (GML) has the mission of acquire,
evaluate and make available long-term records of atmospheric gases[^2].
To achieve that goal, GML gather own and other laboratories
data, releasing observation in a compendium named ObsPack
[@masarie2014obspack]. Specifically, the $CH_4$ ObsPack GLOBALVIEW+
is a comprehensive product consisting in observations from aircrafts, 
ships, surface stations, towers and aircores.
ObsPack include a descriptor named `datasetid` covering:
aircraft-pfp, aircraft-insitu, aircraft-flask, surface-insitu,
surface-flask, surface-pfp, tower-insitu, aircore, shipboard-insitu,
and shipboard-flask. ObsPack product generally contains hundreds of files, each of which 
has different sampling frequencies, hours, and attributes.
It takes time and effort to develop tools to read and process 
each ObsPack product and select observations of interest for specific 
modeling and data analysis purposes.

The NOAA ObsPack data is delivered to the public as NetCDF and 
text files [@masarie2014obspack]. The structure of the files including descriptor fields
depend on the type of file. For instance, the metadata
from aircrafts is different than surface stations, but all the files
include concentrations and other critical fields. 
Given the complexity of ObsPack format, reading and analyzing the data
can be cumbersome. The `rtorf` package provides the 
GHG science and research community a transparent and efficient tool to 
process ObsPack products for GHG modeling and analyses. 
In this manuscript we present `rtorf`, an R package to read,
process and plot NOAA ObsPack data, a software useful and needed
for the community [@R]. For this release, we are 
focused on the $CH_4$ ObsPack GLOBALVIEW+ product.
The general process consists in creating a 
summary of the ObsPack files, reading them in an iteration process, 
filtering, and generating another output and plots.

# Installation

To install `rtorf`, the user must have installed the R package 
remotes and run the following script. This process will install 
all the required dependencies, such as data.table, `cptcity`, an R 
package with more than 7000 color palettes, and `lubridate`, a package 
to manage time and dates [@lu;@cpt]. Then, we call the libraries
to load the function into the environment. `rtorf` is hosted at GitHub,
which allows the implementation of checking the package installation in a 
variety OS.

```{r install, eval = F}
remotes::install_github("noaa-gml/rtorf")
library(rtorf)
```

# Overview

`rtorf` is a collection of function organized together to read and 
process ObsPack files. The general process 
consists in create a 
summary of the ObsPack files, reading them in an iteration process, 
filter and generating another output. As $CH_4$ ObsPACKGLOBALViewplus 5.1,
the product used in this manuscript, includes dataid, we produced
a guide for each of of them available at https://noaa-gml.github.io/rtorf.
Then, in this manuscript we present the processing of `aircraft-insitu`.
The obspack product in this case is `obspack_ch4_1_GLOBALVIEWplus_v5.1_2023-03-08`.

## Summary

We first call the libraries `rtorf` and `data.table`.  Most of objects returned
by `rtorf` are of class `data.table`. Then, we define the datasetid to be 
identified in the name of the files inside the directory with data. 
This process print a summary of the data and if the logical argument verbose
is `TRUE`, print the file being read in each iteration, default is `FALSE`.

```{r}
library(rtorf)
library(data.table)
cate = c("aircraft-pfp",
         "aircraft-insitu",
         "aircraft-flask",
         "surface-insitu",
         "surface-flask", 
         "surface-pfp",   
         "tower-insitu",  
         "aircore",       
         "shipboard-insitu",
         "shipboard-flask")

obs <- "Z:/torf/obspack_ch4_1_GLOBALVIEWplus_v5.1_2023-03-08/data/nc/"
index <- obs_summary(obs = obs, 
                     categories = cate)
```

Once the index of file is built, we can read each file. As we are directing to the
nc directory in ObsPack, with NetCDF files inside, we use the function 
`obs_read_nc`. This function dumps the NetCDF information into a data.table 
with `long` format [@Silge2016]. As the global attributes is attributes in
the NetCDF would result in a data.table with too many columns, we used
the argument `att` equals to FALSE, which is default.
In ground-based datasetid the `solar_time` array is available.
This is useful to select specific observations, for more information check
the site documentation. At the moment, this array is not available for aircraft
observations, hence we select `FALSE`. In this case, we select `verbose` equal
to `TRUE`, to see the name of the files being read.

```{r}
datasetid <- "aircraft-insitu"
df <- obs_read_nc(index = index,
                  categories = datasetid,
                  att = FALSE,
                  solar_time = FALSE,
                  verbose = TRUE)
```

The resulting `data.table` contains `r ncol(df)` columns and 
`r nrow(df)` observations. 
Furthermore, the size of `data.table` is `r format(object.size(df), unit = "Gb")`.
The data includes observations between `r range(df$year)[1]` and 
`r range(df$year)[2]`. Now we can define some parameters to filter our data,
like the year 2020 and spatially data below 8000 meters above sea level (masl) and
focused over north America.

```{r}
df <- df[year == 2020 &
           altitude_final < 8000 &
           latitude < 80 &
           latitude > 10 &
           longitude < -50 &
           longitude > -170]
```

Now we have a `data.table` that contains `r ncol(df)` columns and 
`r nrow(df)` observations. The size of `data.table` is 
`r format(object.size(df), unit = "Gb")`. We can add a column
of time in format "POSIXct" and cut the seconds every 20 seconds.
Usually, aircraft observations every 1 second. Then,
We can simplify the data by calculating averages every 20 seconds.
We perform this task by cutting time very 20 seconds. Then,
we add a new column with the mandatory name `key_time`, which 
will be used to aggregate data with "POSIXct" class, but every 20 seconds.

```{r}
df <- obs_addtime(df)
df$sec2 <- obs_freq(x = df$second,
                    freq = seq(0, 59, 20))
df$key_time <- ISOdatetime(year = df$year,
                           month = df$month,
                           day = df$day,
                           hour = df$hour,
                           min = df$minute,
                           sec = df$sec2,
                           tz = "UTC")
df[1, c("timeUTC", "key_time")]
```

now we can aggregate the data using the function `obs_agg`. The argument cols
indicate which columns will be averaged. Then, we add local time with the 
function `obs_addltime` and we re order the data.table.

```{r}
df2 <- obs_agg(df, 
               cols =  c("year", "month", "day", "hour", "minute", "second",
                         "time", "time_decimal",  "value",
                         "latitude", "longitude", "altitude_final",
                         "pressure", "u", "v", "temperature", "type_altitude"))
df3 <- obs_addltime(df2)
setorderv(df3, cols = c("site_code", "timeUTC"),
          order = c(-1, 1))
```

## Solar or local time

Identifying the local time is important for atmospheric reasons. Sometimes
we need observations when the Planetary Boundary Layer is high,
so that the concentrations are well distributed, generaly 
around 2:00pm. In `rtorf` we use an hierarchical approach
based on the availability of critical information.
Basically, if the solar time array is available, we use
the functon `obs_addstime`. In the begative case
`rtorf` searches for the metadata `site_utc2lst` ot convert
UTC time to local. Finally, in the absence of the mentioned data,
we calculate an approximation of the local time
using the geographical coordinates, as :

$$
lt = UTC + longitude/15 * 60 * 60
$$
Where $lt$ is the local time, $UTC$ the time,
$longitude$ the coordinate.

## Plots

Now we have the data processed and ready to be exported. `rtorf` includes 
a number of functions to save the data as takbulated format in text, csv and 
CSVY[^1] are csv files with a YAML header. This funcitons can be seen in the documentation.
In this last part of the manuscript we will show some visualizations. We included
a function named `obs_plot` which plots data in long format using R base functions.
Here we see data for the month of March 2020. This usefull function allows to plot several
sites and prints the x-axis range.

```{r, fig.width=5, fig.height=3}
obs_plot(df3[month == 3], time = "timeUTC", yfactor = 1e9, type = "b",
         xlab = "UTC time", ylab = expression(CH[4]~ppb))

```
Finally, we show some vertical profiles for the months of January and March of 2020.
We can see how during March of 2020 methane concentrations are lower than
January. This may be due the implementation of Lockdowns [@espinosa2023covid].
A manuscript focused on the impact of COVID-19 on methane emissions
will be submitted soon.

```{r}
x <- df3
x$ch4 <- x$value*1e+9
obs_plot(x, 
         time = "ch4", 
         y = "altitude_final", 
         colu = "month", 
         type = "b", 
         xlab = expression(CH[4]~ppb), 
         ylab = "altitude (m)")
```

## Future work

We are currently porting `rtorf` to python into a package named `pytorf`
with the same functionalities. Furthermore, we expect to implement
WebAsembly with Web-R[^2] in R and Python to run `rtorf` on the browser [@webr].

[^1]: https://csvy.org/
[^2]: https://docs.r-wasm.org/webr/latest/

# Acknowledgements

This project is funded by the NOAA Climate Program Office AC4 and COM programs (NA21OAR4310233 / NA21OAR4310234). This research was supported by NOAA cooperative agreement NA22OAR4320151. Also, thanks to Arlyn Andrews, 
John Miller, Kenneth Schuldt, Kirk Thoning and Andy Jacobson from NOAA GML.

# References
